---
title: "decorate_network"
author: "DLS"
date: "2024-06-02"
output: html_document
---

# There are a couple of different goals here
# First and foremost, we want to make a pretty network
# We will do this by taking the c1.ntw object created by Vcontact and adding usefull info to it (mostly via for loops)
  # First, we want to add host prediction to the network
    # Specifically...
    # 1) If we successfully assigned a host to a phage using CRISPR, we will use the phage_host dataset to assign host taxonomy in the network
    # 2) If a phage does not have a predicted CRISPR host, we will bin its host as "unassigned"
# Load and clean data
```{r}
# Making a pretty network
# Read in network
no_ref = read.table("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/network/no_ref/c1.ntw",
         header = F)

# Breifly, we want to add a column that tells what study a particular phage came from 
library(dplyr)
no_ref = no_ref %>%
  mutate(Study = ifelse(grepl("rosso", no_ref$V1), "bonilla-rosso", 
                        ifelse(grepl("deboutte", no_ref$V1), "deboutte", 
                               ifelse(grepl("busby", no_ref$V1), "busby", "sbardellati"))))
# Fix names
colnames(no_ref) = c("Source", "Target", "Weight", "Study")
no_ref$Source = gsub(pattern = "annotate_", "", x = no_ref$Source)
no_ref$Target = gsub(pattern = "annotate_", "", x = no_ref$Target)
```

# Add Host assignments, taxonomy, and clusters
```{r}
# Load additional data
## host prediction
phage_host = read.csv("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/crispr/all_phage_host_table.csv",
                      header = T)
## Taxonomy prediction
phage_tax = read.csv("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/taxonomy/viral_taxonomy.csv", header = T, row.names = 1)
## Cluster assignment
clusters = read.csv("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/network/no_ref/genome_by_genome_overview.csv")
clusters$Genome = gsub("annotate_", "", x = clusters$Genome)
clusters_clean = clusters %>% 
  dplyr::select(Genome, VC)

# Add data to network object
no_ref_host_tax_clust = no_ref %>%
  left_join(phage_host, join_by(Source == phage)) %>%
  left_join(phage_tax, join_by(Source == vOTU)) %>%
  left_join(clusters_clean, join_by(Source == Genome))
```

# Add an aditional column that is Host_study
## This will allow us to "grey out" previously described phage in our eventual network
```{r}
final_df = no_ref_host_tax_clust %>%
  mutate(host_study = paste(host, Study, sep = "_"))
```

# Export df
```{r}
# export it
# write.csv(final_df, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/network/network_object.csv", row.names = F, quote = F)
```

# Now lets try and improve our assignemnts using clusters predicted by vContact
## If a phage with an unassigned host is in a cluster with phage that have assigned host, give the unassigned phage the assigned host
    # We will only do this for unambiguous clusters (i.e. clusters that only have one assigned host)
    # If a cluster contains phage with more than one predicted host, all phage will retain their original host predictions + unassigned phage will remain unassigned
# Update proposed hosts using cluster info
```{r}
# Now lets use another loop to classify the host of phage which are unassigned 
# If an unassigned phage clusters with an assigned one, give it the assigned's host
line = c()
clust = c()
all_clust = c()
clust_hosts = c()
new_host = c()
out = c()
# for testing clust = "VC_7_0"
for (i in 1:nrow(final_df)) {
  # Go line by line through the df
  line = final_df[i,]
  # and pull each cluster from the line
  clust = line$VC
    if (clust == "") { # if the phage in question is a singleton
      new_host = line$host # dont change its host
    } else {
  # then subset the whole df to just that cluster
  all_clust = final_df[final_df$VC == as.character(clust),]
  # and pull out the unique host names
  clust_hosts = unique(all_clust$host)
  # next, a series of if else statements:
  # If there is an unassigned host in this VC... 
  if (NA %in% clust_hosts) {
    # And if there is only the unnassigned host
    if (length(clust_hosts) == 1) {
      # then make the "new" host variable = to NA
      new_host = NA
      # Otherwise...
    } else{
      # Drop the "NA" variable from the list of hosts in the cluster
      new_host = clust_hosts[-which(is.na(clust_hosts))]
      # Then, if there is more than 1 host in the list of hosts...
      if (length(new_host) > 1) {
        # Keep the old assigned host for that line (this is a cluster with more than one host in it)
        new_host = line$host
        # Otherwise (only one host remains in the list of hosts)....
      } else {
        # Assign the remaining host 
        new_host = new_host
      }
    }
  } else{
    new_host = line$host
  }
    }
  out = c(out, new_host)
}
# Add a new host col based on the results of this loop
updated_host = final_df
updated_host$host_2 = out
```
# check our work
```{r}
# After using clusters to assign hosts, are there any clusters with more than one host assigned to them?
# Select just the important cols
updated_host_sub = updated_host %>%
  dplyr::select(Source, host, host_2, VC)
# Drop singletons (so they are not interpreted as a cluster)
updated_host_sub_no_sing = updated_host_sub[!updated_host_sub$VC == "",]
# aggregate and look at result
# VC = name of viral cluster, Count = number of hosts present
updated_host_sub_no_sing %>%
  dplyr::group_by(VC) %>% 
  dplyr::summarise(count = n_distinct(host_2)) %>%
  dplyr::arrange(desc(count))
# two cluster have more than one host in them!
# VC_4_0 and VC_254_0 have more than one host assigned

# Take a peak at these
# VC_4_0
VC_4_0 = updated_host_sub_no_sing[updated_host_sub_no_sing$VC == "VC_4_0",]
unique(VC_4_0$Source)
unique(VC_4_0$host_2)
# 5 different phage in cluster
# Contains phage from our study and from busby
# Target unassigned"  "Gilliamella" "Serratia" 


# VC_254_0
VC_254_0 = updated_host_sub_no_sing[updated_host_sub_no_sing$VC == "VC_254_0",]
unique(VC_254_0$Source)
unique(VC_254_0$host_2)
# 2 phage in cluster
# Contains only phage from our study
# Target Bifidobacterium" "Lactobacillus"  

# Two clusters are "ambiguous" -- Doesn't seem that bad 
```

# Repeat process, but with phage taxonomy
```{r}
# set Family == NA to Family == ""
updated_host_fam_fix = updated_host %>%
  mutate(Family_fix = ifelse(is.na(Family), "", Family))

# Now lets use another loop to classify the host of phage which are unassigned 
# If an unassigned phage clusters with an assigned one, give it the assigned's host
line = c()
clust = c()
all_clust = c()
clust_fam = c()
new_fam = c()
out = c()
# for testing clust = "VC_7_0"
for (i in 1:nrow(updated_host_fam_fix)) {
  # Go line by line through the df
  line = updated_host_fam_fix[i,]
  # and pull each cluster from the line
  clust = line$VC
    if (clust == "") { # if the phage in question is a singleton
      new_fam = line$Family_fix # dont change its host
    } else {
  # then subset the whole df to just that cluster
  all_clust = updated_host_fam_fix[updated_host_fam_fix$VC == as.character(clust),]
  # and pull out the unique host names
  clust_fam = unique(all_clust$Family_fix)
  # next, a series of if else statements:
  # If there is an unassigned host in this VC... 
  if ("" %in% clust_fam) {
    # And if there is only the unnassigned host
    if (length(clust_fam) == 1) {
      # then make the "new" host variable = to ""
      new_fam = ""
      # Otherwise...
    } else{
      # Drop the "NA" variable from the list of hosts in the cluster
      new_fam = clust_fam[-which(clust_fam == "")]
      # Then, if there is more than 1 host in the list of hosts...
      if (length(new_fam) > 1) {
        # Keep the old assigned host for that line (this is a cluster with more than one host in it)
        new_fam = line$Family_fix
        # Otherwise (only one host remains in the list of hosts)....
      } else {
        # Assign the remaining host 
        new_fam = new_fam
      }
    }
  } else{
    new_fam = line$Family_fix
  }
    }
  out = c(out, new_fam)
}
# Add a new host col based on the results of this loop
updated_host_tax = updated_host_fam_fix
updated_host_tax$Family_2 = out

```
# check our work
```{r}
# After using clusters to assign hosts, are there any clusters with more than one host assigned to them?
# Select just the important cols
updated_host_tax_sub = updated_host_tax %>%
  dplyr::select(Source, Family_fix, Family_2, VC)
# Drop singletons (so they are not interpreted as a cluster)
updated_host_tax_sub_no_sing = updated_host_tax_sub[!updated_host_tax_sub$VC == "",]
# aggregate and look at result
# VC = name of viral cluster, Count = number of hosts present
updated_host_tax_sub_no_sing %>%
  dplyr::group_by(VC) %>% 
  dplyr::summarise(count = n_distinct(Family_2)) %>%
  dplyr::arrange(desc(count))
# No ambiguous taxonomy clusters
```

# Clean and export dfs
```{r}
# Select just the cols we want
cleaned_final = updated_host_tax %>%
  dplyr::select(Source, Target, Weight, Study, VC, Class, Family_2, host_2) %>%
  mutate(Family = Family_2) %>%
  mutate(Host = host_2) %>%
  mutate(Host_study = paste(host_2, Study, sep = " ")) %>%
  dplyr::select(Source, Target, Weight, Study, VC, Class, Family, Host, Host_study)
# export it
# write.csv(cleaned_final, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/network/network_object_clustered.csv", row.names = F, quote = F)

updated_phage_host_tax = cleaned_final %>%
  mutate(Phage = Source) %>%
  dplyr::select(Phage, Host, Family) %>%
  unique() %>%
  replace_na(list(Host = "", Family = ""))

# export it
# write.csv(updated_phage_host_tax, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/Phage_host_analysis/updated_host_and_taxonomy.csv", na = "")
```

# Lets quickly check out our clusters to try and find one to run through clinker
```{r}
cleaned_final %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Study)) %>%
  arrange(desc(n)) 

# These clusters contain phage from all studies
# "VC_138_0"     4
# "VC_13_0"      4
# "VC_149_0"     4
# "VC_18_0"      4
# "VC_22_0"      4
# "VC_26_0"      4
# "VC_35_0"      4
# "VC_85_0"      4

# VC_138
cleaned_final %>% 
  filter(VC == "VC_138_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 11 vOTUs
VC_138 = cleaned_final %>% 
  filter(VC == "VC_138_0") %>%
  select(Source) %>%
  unique()
write.table(VC_138, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/clinker/VC_138.csv", sep = ",", col.names = F, quote = F, row.names = F)

# VC_13
cleaned_final %>% 
  filter(VC == "VC_13_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 32 vOTUs

# VC_149
cleaned_final %>% 
  filter(VC == "VC_149_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 9 vOTUs
VC_149 = cleaned_final %>% 
  filter(VC == "VC_149_0") %>%
  select(Source) %>%
  unique()
write.table(VC_149, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/clinker/VC_149.csv", sep = ",", col.names = F, quote = F, row.names = F)

# VC_18
cleaned_final %>% 
  filter(VC == "VC_18_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 34 vOTUs
VC_18 = cleaned_final %>% 
  filter(VC == "VC_18_0") %>%
  select(Source) %>%
  unique()
write.table(VC_18, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/clinker/VC_18.csv", sep = ",", col.names = F, quote = F, row.names = F)

# VC_22_0
cleaned_final %>% 
  filter(VC == "VC_22_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 6 vOTUs
VC_22 = cleaned_final %>% 
  filter(VC == "VC_22_0") %>%
  select(Source) %>%
  unique()
write.table(VC_22, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/clinker/VC_22.csv", sep = ",", col.names = F, quote = F, row.names = F)

# VC_35_0
cleaned_final %>% 
  filter(VC == "VC_35_0") %>%
  dplyr::group_by(VC) %>%
  dplyr::summarise(n=n_distinct(Source))
# has 14 vOTUs
VC_35 = cleaned_final %>% 
  filter(VC == "VC_35_0") %>%
  select(Source) %>%
  unique()
write.table(VC_35, "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/clinker/VC_35.csv", sep = ",", col.names = F, quote = F, row.names = F)
```