---
title: "Build_vOTU_table"
author: "DLS"
date: "2024-05-28"
output: html_document
---

# The goal of this script is to take the vOTU coverage values we calculated previously and construct a vOTU table.
# This script will also take into account reviewer feedback to change what phage we consider present/absent
# In particular, we... 
  ## Drop vOTUs if they have a horizontal coverage of <50% 
  ## Build that vOTU table
  ## Normalize coverage counts to "coverage per million reads"
  ## Drop phage which are present at <1 

# Load packages
```{r}
library(tidyverse)
library(vegan)
library(lme4)
library(lmerTest)
library(reshape2)
```

# Set wd and load data in
```{r}
# Set path
setwd("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/coverM_out")
# Make file obj
files = list.files(path = "./", pattern = "*.tsv")
# Extract phage names from first col of first file
phage = read.table(files[1], header=FALSE, sep="\t")[,1]
# Clean names and make new df
phage_clean = phage[-1]
```
# Filter by "horizontal coverage"
## i.e. only retain phage which have at least 1/2 their genome covered by a read
```{r}
# Make function for building vOTU table with horizontal coverage as the value
fn = function(x) {
  y = read.table(x, header = F, sep = "\t")
  name = paste(y[1,3])
  name <- gsub(".fna Mean", "", name)
  z = data.frame(name1 = as.numeric(y[-1,3]))
  names(z) = name
  return(z)
}
# Run that function
hor_cov_values_list = apply(X = matrix(files), FUN = fn, MARGIN = 1)
hor_cov_values_df = do.call(what = cbind, args = hor_cov_values_list)
# Check if there are any phage with a min covered fraction less than 0.5
hor_cov_values_df %>% 
  melt() %>% # melt the df
  filter(value > 0) %>% # subset to just non-0 covered fractions 
  filter(value < 0.5) # now filter down to those with less than 0.5 min covered fraction
# This produces an empty df. So when a phage is present in a sample, it has a horzontal coverage of at least 0.5
# For reference, we can filter down to a horizontal coverage of 0.75 and still not drop any
## This means, after mapping our reads back to vOTU genomes, all genomes which recruit reads recruit enough reads to cover at least 75% of the genome
# Lets move forward!
```
# Build vOTU table 
```{r}
# Make function for building vOTU table with trimmed mean as the value
fn = function(x) {
  y = read.table(x, header = F, sep = "\t")
  name = paste(y[1,2])
  name <- gsub(".fna Mean", "", name)
  z = data.frame(name1 = as.numeric(y[-1,2]))
  names(z) = name
  return(z)
}

# Run that function
trimmed_mean_values_list = apply(X = matrix(files), FUN = fn, MARGIN = 1)
trimmed_mean_values_df = do.call(what = cbind, args = trimmed_mean_values_list)
# Correct the name of phage
name = str_split(string = files, pattern = "_", n = 4, simplify = T)
name_clean = paste(name[,2], name[,3], name[,1], sep = "_")
colnames(trimmed_mean_values_df) = name_clean

# Build the vOTU table
phage_df = data.frame("genome" = phage_clean, trimmed_mean_values_df)
rownames(phage_df) = phage_df$genome
phage_df = data.frame(t(phage_df[-1]))
saveRDS(object = phage_df, 
        file = "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/coverage_table/5kb_phage.RDS")
```
# Normalize to coverage per millio reads
```{r}
# vOTU table
raw_vOTU_table = readRDS("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/coverage_table/5kb_phage.RDS")
# meta data table
meta = read.csv("/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/coverage_table/Bee_phage_meta.csv", header = T)
# Drop note section cause its long, ugly, and we dont need it
meta = meta[!colnames(meta) == "Notes"]
# Subset to just virome samples
meta_virome = meta[meta$Sample %in% rownames(raw_vOTU_table),]

# Normalize phage counts
# Reorder phage df to match metadata
phage_df_order = raw_vOTU_table[order(match(rownames(raw_vOTU_table), meta_virome$Sample)),]
# Double check that this worked
# perform normalization calc
phage_df_norm = (phage_df_order/meta_virome$filtered_read_count)*1000000
# Abundance of phage is now expressed as "number of phage sequences per 1 million reads"
# i.e. The number of times we were able to constuct a particular phage sequence given 1 million reads
```
# Apply abundance cut off
```{r}
# now filter to only the most abundant phage, we will exclude phage with a coverage of less than 1
phage_df_norm[phage_df_norm < 1] = 0
# Now we drop newly absenet phage
phage_df_no_0 = phage_df_norm[,!colSums(phage_df_norm) == 0]
# Left with 490 ~active~ phage
saveRDS(object = phage_df_no_0, file = "/Users/dinosbardellati/Desktop/PhD_stuff/projects/Bee_phage/5kb/reviewer_comments/coverage_table/5kb_phage_normalized.RDS")
```
